---
title: "Employee Absenteeism"
author: "Nithya Vembu"
date: "2/15/2021"
output: 
  html_document:
    keep_md: yes
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing Required Libraries 
```{r message=FALSE, warning=FALSE, error=FALSE}
#install.packages("tidyverse", repos = "http://cran.us.r-project.org")
library(tidyverse)
#install.packages("ggplot2", repos = "http://cran.us.r-project.org")
library(ggplot2)
#install.packages("VIM", repos = "http://cran.us.r-project.org")
library(VIM)
#install.packages("plyr", repos = "http://cran.us.r-project.org")
library(plyr)
#install.packages("dplyr", repos = "http://cran.us.r-project.org")
#library(dplyr)
#install.packages("lubridate", repos = "http://cran.us.r-project.org")
library(lubridate)
#install.packages("constants", repos = "http://cran.us.r-project.org")
library(constants)
#install.packages("corrplot", repos = "http://cran.us.r-project.org")
library(corrplot)
#install.packages("rpart", repos = "http://cran.us.r-project.org")
library(rpart)
#install.packages("rpart.plot", repos = "http://cran.us.r-project.org")
library(rpart.plot)
#install.packages("e1071", repos = "http://cran.us.r-project.org")
library(e1071)
#install.packages("devtools", repos = "http://cran.us.r-project.org")
library(devtools) 
#install.packages("caret", repos = "http://cran.us.r-project.org")
library(caret)
install_github("vqv/ggbiplot")
library(ggbiplot)
#install.packages("randomForest", repos = "http://cran.us.r-project.org")
library(randomForest)
#install.packages("gridExtra", repos = "http://cran.us.r-project.org")
library(gridExtra)
```


## Importing Data Files
```{r}
train_data <- read.table("/Users/nithyavembu/Downloads/Absenteeism_at_work_train.csv", header=TRUE, sep=",")
test_data <- read.table("/Users/nithyavembu/Downloads/Absenteeism_at_work.csv", header=TRUE, sep = ";")
```

##Exploratory Data Analysis - Part I
**1. Checking variable types in _train_data_ and _test_data_**
```{r}
str(train_data)
str(test_data)
```
**2. Observation**<br/>
1. _train_data_ has two attributes as _factors_ : _Age_ and _Work.load.Average.day_<br/>
2. Attribute _work.load.Average.day_  in _test_data_  has a different name.<br/> 

**3. Identified unknown levels "0" and "R" in attribute _Age_ in _train_data_**
```{r}
levels_train_data_age <- levels(train_data$Age) 
print(levels_train_data_age)
# Counting number of samples with "0" and "R"
#count(train_data$Age == "R")
#count(train_data$Age == "0")
# Removing samples with NA values train_data
train_data <- train_data[!train_data$Age %in% c("0","R"),]
```

**4. Converting attribute _Age_ from type _factor_ to _number_** 
```{r}
train_data$Age <- as.integer(as.character(train_data$Age))
```

**5. Rename attribute _work.load.Average.day_ in _test_data_**
```{r}
names(test_data)[names(test_data)=="Work.load.Average.day."] <- "Work.load.Average.day"
```

**6. Identified different variable format of attrbute _work.load.Average.day_ in _train_data_ and _test_data_ and converted them to a uniform format**<br/>
```{r}
train_data$Work.load.Average.day <- as.numeric(gsub(",", ".", gsub("\\.", "", train_data$Work.load.Average.day)))
```

##Handling Missing Values
```{r}
# Check for NA values
sapply(train_data, function(x) sum(is.na(x)))
```
```{r}
# Remove rows with NA values
train_data_clean <- na.omit(train_data)
```

## Grouping target variable
```{r}
#Grouping target variable in Training Set
train_data_clean$Absenteeism.Hours.Group[train_data_clean$Absenteeism.time.in.hours == 0] <- "0"
train_data_clean$Absenteeism.Hours.Group[train_data_clean$Absenteeism.time.in.hours == 1] <- "1"
train_data_clean$Absenteeism.Hours.Group[train_data_clean$Absenteeism.time.in.hours == 2] <- "2"
train_data_clean$Absenteeism.Hours.Group[train_data_clean$Absenteeism.time.in.hours == 3] <- "3"
train_data_clean$Absenteeism.Hours.Group[train_data_clean$Absenteeism.time.in.hours >=4 &  train_data_clean$Absenteeism.time.in.hours <=7] <- "4"
train_data_clean$Absenteeism.Hours.Group[train_data_clean$Absenteeism.time.in.hours == 8] <- "5"
train_data_clean$Absenteeism.Hours.Group[train_data_clean$Absenteeism.time.in.hours >= 9] <- "6"
```

```{r}
# Grouping target variable in Testing Set
test_data$Absenteeism.Hours.Group[test_data$Absenteeism.time.in.hours == 0] <- "0"
test_data$Absenteeism.Hours.Group[test_data$Absenteeism.time.in.hours == 1] <- "1"
test_data$Absenteeism.Hours.Group[test_data$Absenteeism.time.in.hours == 2] <- "2"
test_data$Absenteeism.Hours.Group[test_data$Absenteeism.time.in.hours == 3] <- "3"
test_data$Absenteeism.Hours.Group[test_data$Absenteeism.time.in.hours >=4 &  test_data$Absenteeism.time.in.hours <=7] <- "4"
test_data$Absenteeism.Hours.Group[test_data$Absenteeism.time.in.hours == 8] <- "5"
test_data$Absenteeism.Hours.Group[test_data$Absenteeism.time.in.hours >= 9] <- "6"
```

```{r}
#Dropping Absenteeism from Train and Test Data
train_data_clean$Absenteeism.time.in.hours <- NULL
test_data$Absenteeism.time.in.hours <- NULL
```

```{r}
# Converting Target Variable as Factors
train_data_clean$Absenteeism.Hours.Group <- as.factor(train_data_clean$Absenteeism.Hours.Group)
test_data$Absenteeism.Hours.Group <- as.factor(test_data$Absenteeism.Hours.Group)
```

**Checking var types for cleaned testing and training data set**
```{r}
str(train_data_clean)
str(test_data)
```


## Exploratory Data Analysis - Part II
**1. Corelation Plot**
```{r}
# temporary variable train_corr
train_corr <- train_data_clean
# change target group into numeric
train_corr$Absenteeism.Hours.Group <- as.numeric(train_corr$Absenteeism.Hours.Group)
# plot
corrplot(cor(train_corr), order = "hclust", tl.col = 'black', tl.cex = .75)
```
<br/>**Observation from Corelation Plot**
 <br/>1. variables _Weight_ and _Body.Mass.Index_ are highly corelated
 <br/>2. variables _Service.time_ and _Age_ are highly corelated

**<br/>2. Absentees by Month and Day of Week**
```{r}
# 1. Absentees by Month
x <- train_data_clean[c("Month.of.absence", "Day.of.the.week")]
x$Month.of.absence <- as.factor(x$Month.of.absence)
x <- x %>%
  mutate(Month.of.absence= fct_recode(Month.of.absence,'None'='0','Jan'='1','Feb'='2','Mar'='3','Apr'='4','May'='5',
                                      'Jun'='6','Jul'='7','Aug'='8','Sep'='9','Oct'='10','Nov'='11','Dec'='12') )

x$Day.of.the.week <- as.factor(x$Day.of.the.week)
x <- x %>%
  mutate(Day.of.the.week = fct_recode(Day.of.the.week,"Monday"="2","Tuesday"="3","Wednesday"="4","Thursday"="5","Friday"="6"))
g <- ggplot(x, aes(Month.of.absence))
g + geom_bar(aes(fill=Day.of.the.week), width = 0.5) + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) + 
  labs(x = "Month" ,y = "#Absentees", title= "Absenteeism across a Month and it's Day of the week")

```
**<br/>3. Absentees by Season**
```{r}
# 2. Absentees by Season
x <- train_data_clean[c("Seasons")]
x$Seasons <- as.factor(x$Seasons)
x <- x %>%
  mutate(Seasons= fct_recode(Seasons,'Summer'='1','Autumn'='2','Winter'='3','Spring'='4'))
ggplot(x, aes(x =   Seasons,fill = Seasons)) + geom_bar() + 
  labs(x = "Season" ,y = "#Absentees", title= "Absenteeism across Seasons")
```
**<br/>4. Absentees by Reasons**

```{r}
x <- count(train_data_clean, "Reason.for.absence")
x$Reason.for.absence <- as.factor(x$Reason.for.absence)
x <- x %>%
  mutate(Reason.for.absence = fct_recode(Reason.for.absence,'Infectious,parasitic diseases'='0',
                                         'Neoplasms'='1','Diseases of the blood'='2','Endocrine and metabolic diseases'='3','Mental and behavioural disorders'='4', 
                                         'Diseases of the nervous system'='5','Diseases of the eye and adnexa'='6','Diseases of the ear and mastoid process'='7',
                                         'Diseases of the circulatory system'='8','Diseases of the respiratory system'='9','Diseases of the digestive system'='10', 
                                         'Diseases of the skin and subcutaneous tissue'='11','Diseases of the musculoskeletal system and connective tissue'='12', 
                                         'Diseases of the genitourinary system'='13','Pregnancy, childbirth and the puerperium'='14','Certain conditions originating in the perinatal'='15', 
                                         'Congenital malformations, deformations and chromosomal abnormalities'= '16','Symptoms, signs and abnormal clinical  findings'='17',
                                         'Injury, poisoning and certain other consequences of external causes'= '18','causes of morbidity and mortality'='19',
                                         'Factors influencing health status and contact with health services'='21','patient follow-up'='22','medical consultation'='23','blood donation'='24',
                                         'laboratory examination'='25','unjustified absence'='26','physiotherapy'='27','dental consultation'='28'))

ggplot(x,aes(x = Reason.for.absence, y= freq, fill= Reason.for.absence)) + geom_bar(stat = 'identity') + coord_flip() + theme(legend.position='none') +  
  xlab('Reason for absence') + ylab('#Absentees')
```
**<br/>5. Absentees by Number of Pets**
```{r}
x <- train_data_clean[c("Pet")]
x$Pet <- as.factor(x$Pet)
g = ggplot(x, aes(Pet))
g + geom_bar(aes(fill=Pet), width = 0.6) + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) + 
  labs(x = "#Pets" ,y = "#Absentees", title= "Absenteeism vs  #Pets")
```
**<br/>6. Absentees by Number of Children**
```{r}
x <- train_data_clean[c("Son")]
x$Son <- as.factor(x$Son)
g = ggplot(x, aes(Son))
g + geom_bar(aes(fill=Son), width = 0.6) + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) + 
  labs(x = "#Children" ,y = "#Absentees", title= "Absenteeism vs #Children")
```

**<br/>7. Absentees by Social Drinking and Smoking**
```{r}
x <- train_data_clean[c("Social.drinker")]
x$Social.drinker <- as.factor(x$Social.drinker)
x <- x %>%
  mutate(Social.drinker= fct_recode(Social.drinker,'No'='0','Yes'='1') )
g = ggplot(x, aes(train_data_clean$Social.drinker))
g <- g + geom_bar(aes(fill=Social.drinker), width = 0.6) + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) + 
  labs(x = "Social Drinker" ,y = "#Absentees", title= "Absenteeism vs Social Drinking")

x <- train_data_clean[c("Social.smoker")]
x$Social.smoker <- as.factor(x$Social.smoker)
x <- x %>%
  mutate(Social.smoker= fct_recode(Social.smoker,'No'='0','Yes'='1') )
g1 = ggplot(x, aes(Social.smoker))
g1 <- g1 + geom_bar(aes(fill=Social.smoker), width = 0.6) + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) + 
  labs(x = "Social Smoker" ,y = "#Absentees", title= "Absenteeism vs Social Smoking")
grid.arrange(g,g1, nrow = 1)
```

**<br/>8. Absentees by Hit Target**
```{r}
x <- train_data_clean[c("Hit.target")]
x$Hit.target <- as.factor(x$Hit.target)
g = ggplot(x, aes(Hit.target))
g + geom_bar(aes(fill=Hit.target), width = 0.6) + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) + 
  labs(x = "Hit Target Periodically" ,y = "#Absentees", title= "Absenteeism vs %Target Achievement")
```

**<br/>9. Absentees by Disciplinary Action**
```{r}
x <- train_data_clean[c("Disciplinary.failure")]
x$Disciplinary.failure <- as.factor(x$Disciplinary.failure)
x <- x %>%
  mutate(Disciplinary.failure= fct_recode(Disciplinary.failure,'No'='0','Yes'='1') )
g = ggplot(x, aes(Disciplinary.failure))
g + geom_bar(aes(fill=Disciplinary.failure), width = 0.6) + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) + 
  labs(x = "Disciplinary Action" ,y = "#Absentees", title= "Absenteeism vs Disciplinary Action")
```

**<br/>10. Absentees by Education**
```{r}
x <- train_data_clean[c("Education")]
x$Education <- as.factor(x$Education)
x <- x %>%
  mutate(Education= fct_recode(Education,'highschool'='1','graduate'='2','postgraduate'='3','master& doctrate'='4') )
g = ggplot(x, aes(Education))
g + geom_bar(aes(fill=Education), width = 0.6) + 
  theme(axis.text.x = element_text(angle=0, vjust=0.6)) + 
  labs(x = "Education" ,y = "#Absentees", title= "Absenteeism vs Education")

```

## Decision Tree Model
**Basic Model**
```{r}
DT_model <- rpart(train_data_clean$Absenteeism.Hours.Group~., data = train_data_clean, method = 'class')
rpart.plot(DT_model)
predicted_values<-predict(DT_model, test_data, type = 'class')
table_mat <- table(test_data$Absenteeism.Hours.Group, predicted_values)
# Confusion Matrix
print(table_mat)
# Accuracy
accuracy_dt<- sum(diag(table_mat)) / sum(table_mat) * 100
print(accuracy_dt)
```

**Fine Tune Decision Tree Model**
```{r}
accuracy_tune <- function(DT_model) {
      predicted_values <- predict(DT_model, test_data, type = 'class')
      table_mat <- table(test_data$Absenteeism.Hours.Group, predicted_values)
      accuracy_Test <- sum(diag(table_mat)) / sum(table_mat) * 100
      print(table_mat)
      accuracy_Test
   }

control <- rpart.control(minsplit = 2, minbucket = round(5 / 3), maxdepth = 20, cp = 0)
tune_DT_model <- rpart(train_data_clean$Absenteeism.Hours.Group~., data = train_data_clean, method = 'class', control = control)


accuracy_tune(tune_DT_model)
```


## Stochastic Vector Machines
```{r}
# Separating data into x_train, y_train, x_test, y_test
x_train <- train_data_clean[-21]
y_train <- train_data_clean[,21]
x_test <- test_data[-21]
y_test <- test_data[,21]

# SVM model
svm_model <- svm(train_data_clean$Absenteeism.Hours.Group ~ ., data=train_data_clean, kernel="polynomial", gamma=1, cost=1, degree = 5, tolerance = 0.001, cross = 10)
print(svm_model)
table_mat<-table(predict(svm_model, x_test), y_test)
accuracy_test <- sum(diag(table_mat)) / sum(table_mat)*100
print(accuracy_test)
```

## Random Forrest Model
```{r}
rf_model <- randomForest(train_data_clean$Absenteeism.Hours.Group ~ ., data = train_data_clean, ntree = 100, importance=TRUE)
rf_model
table_mat<-table(predict(rf_model, x_test), y_test)
accuracy_test <- sum(diag(table_mat)) / sum(table_mat)*100
print(accuracy_test)
```

## Principal Component Analysis
**1. Calculating PC**
```{r}
x_train.pca <- prcomp(x_train, center=TRUE, scale. = TRUE)
summary(x_train.pca)
```
**<br/>2. Plot PCs againt respective variance**
```{r}
#plot(method) 
screeplot(x_train.pca, type="lines",col=2)
```
**<br/>3. Plot PCs and their respective cumulative proportions**
```{r}
#plot cum proportions
vars <- apply(x_train.pca$x, 2, var) 
props <- vars/sum(vars) 
x_train.pca.vars <- cumsum(props) 
plot(x_train.pca.vars, type="l")
```
**<br/>4. Plot for explained variance**
```{r}
g <- ggbiplot(x_train.pca, obs.scale = 1, var.scale = 1, groups = y_train, ellipse = TRUE, circle = TRUE)
g <- g + scale_color_discrete((name = ''))
g <- g + theme(legend.direction = 'horizontal', legend.position = 'top') 
print(g)
```
**<br/>5. Generated PC for training set and testing set**
```{r}
x_train_trans = preProcess(x_train, method=c("BoxCox", "center", "scale", "pca"))
PC = predict(x_train_trans, x_train)
PC_test = predict(x_train_trans, x_test)
```
**<br/>6. Using PC with 95% cumulative proportion in SVM model**
```{r}
svm_model <- svm(y_train ~ ., data=PC, kernel="polynomial", gamma=1, cost=1, degree=5, tolerance = 0.001, cross = 10)
print(svm_model)
table_mat<-table(predict(svm_model, PC_test), y_test)
accuracy_test <- sum(diag(table_mat)) / sum(table_mat)*100
print(accuracy_test)
```
**<br/>6. Using PC with different number of components in the SVM model**
```{r}

PC2 = PC[1:10]
PC_test2 = PC_test[1:10]
svm_model <- svm(y_train ~ ., data=PC2, kernel="polynomial", gamma=1, cost=1, degree=5, tolerance = 0.001, cross = 10)
print(svm_model)
table_mat2<-table(predict(svm_model, PC_test2), y_test)
accuracy_test2 <- sum(diag(table_mat2)) / sum(table_mat2)*100
print(accuracy_test2)
```


#### Extra
###Correlation
```{r}
library(corrplot)
train_data_clean_corr <- train_data_clean
train_data_clean_corr$Absenteeism.Hours.Group <- as.numeric(train_data_clean_corr$Absenteeism.Hours.Group)
corrplot(cor(train_data_clean_corr), order = "hclust", tl.col = 'black', tl.cex = .75)
```

```{r}
str(train_data_clean_corr)
```

```{r message=FALSE}
train_data_clean_rm_cor <- train_data_clean
train_data_clean_rm_cor$Weight <- NULL
train_data_clean_rm_cor$Service.time <- NULL
train_data_clean_rm_cor$Seasons <- NULL

test_data_clean_rm_cor <- test_data
test_data_clean_rm_cor$Weight <- NULL
test_data_clean_rm_cor$Service.time <- NULL
test_data_clean_rm_cor$Seasons <- NULL

x_train_rm_cor <- train_data_clean_rm_cor[-17]
y_train_rm_cor <- as.data.frame(train_data_clean_rm_cor[,17])
x_test_rm_cor <- test_data_clean_rm_cor[-18]
y_test_rm_cor <- as.data.frame(test_data_clean_rm_cor[,18])

colnames(x_train_rm_cor) <- make.names(colnames(x_train_rm_cor))
names(y_train_rm_cor) <- "Absenteeism.Hours.Group"
colnames(x_test_rm_cor) <- make.names(colnames(x_test_rm_cor))
names(y_test_rm_cor) <- "Absenteeism.Hours.Group"

svm_model <- svm(y_train_rm_cor$Absenteeism.Hours.Group ~ ., data=x_train_rm_cor, kernel="polynomial", gamma=1, cost=1, tolerance = 0.001, cross = 10)
svm_model
table_mat<-table(predict(svm_model, x_test_rm_cor), y_test_rm_cor$Absenteeism.Hours.Group)
accuracy_test <- sum(diag(table_mat)) / sum(table_mat)
accuracy_test
```


```{r}
rf_model <- randomForest(train_data_clean$Absenteeism.Hours.Group ~ ., data = train_data_clean, ntree = 100, importance=TRUE)
rf_model
```

```{r}
round(importance(rf_model), 3)
varImpPlot(rf_model, pch = 20, main = "Importance of Variables")
```

